---
title: "Computer Lab ECFM: Population dynamics and harvesting"
author: "Viktor Thunell, Anna Gårdmark & Max Lindmark"
date: "7 September 2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim
Biological models of fish populations are used to inform national and international governmental institutions of the state of a populations to support decisions on management objectives or allowed catches. The overall aim of the exercise is 1) to understand how biological models can be used to describe growth of fish populations, for setting catch limits in management and 2) to get better feel of the strengths and limitations of using such models.

# Overview
In this computer based exercise you will use a mathematical model (the Schaefer model [1]) to explore how harvesting (i.e. fishing) changes population numbers and growth rates. The Schaefer model is used to compute the maximum sustainable yield (MSY) which is the largest harvest that a fishstock can produce in the long term [3] and a central measure for fisheries management. We use this model to explore population growth, effects of fishing and MSY. 

# Software tool
You will use R to build and explore the model. R is a statistical software that is very popular in especially biological sciences and is Open Source (free to use and modify). **We expect NO prior knowledge of R nor expect you to learn any R-coding in this exercise**. If you haven't used R before, consider this as an opportunity to get a glimpse of what you can do with this software. If you have used R before, then this may be an opportunity to learn more.

R should be available on university computers but if you like to install R and R-studio on your own computer, please refer to the document **"First time R user"**. R-studio is an interface for R that provides a good overview and has many helpful tools. In this document, R-code is text with grey background. You should copy these chunks of code and paste it to an R-script in R-studio (your instructor will show you how). For some, this computer lab might include some repetition of calculus and algebra, but you will also practice some basic population dynamics and equilibrium theory. Try to think about the problems you meet from an ecological perspective, math and programming are the just tools that help us understand ecology.

Ask your instructor when you get stuck!
 
# The Schaefer Model
The Schaefer model is a harvest model based on the logistic growth equation (the Verhulst model). This is a continuous-time model and the differential equation below describes the rate of change in number of individuals, $N$: 
$$dN/dt = rN(1-N/K) - qEN$$ 
where $r$ is the intrinsic growth rate, $K$ is the carrying capacity, $q$ is catchability and $E$ is harvesting effort. In this lab, $N$ is referred to as "density" (biomass or number of individuals of a population in a water body). Below is some R-code to explore the population density over time in the absence of harvesting to understand the logistic growth model without harvesting.

## Part 1: Recap the logistic growth model without harvesting
This part is mainly to become familiar with the growth model (with fishing mortality, $qEN$, excluded) and read in parameters that you will use later. Essentially just copy paste the chunks of code below until you reach Question #1.
```{r}
# Two parameters govern the dynamics of the population: the growth rate (r) and  
# the carrying capacity (K). 
# We assume arbitrary values, it does not really matter here.
r <- 0.1
K <- 40

# We also need to set the initial population density and the time range 
N_ini <- 2
t <- 100

# Now we can calculate the density at the next time step!
N2 <- r * N_ini * (1-(N_ini/K)) + N_ini
N3 <- r * N2 * (1-(N2/K)) + N2
#... etc
```

We could iterate this procedure manually, but it would be extremely tedious. This is where the loop comes in handy, which is used in programming to automate tasks. Basically, if you find yourself copy-pasting code like this, you can automate it. Here we will use a for loop. This lab is not about writing loops but they are just introduced so that you can get a feel of (1) how it is done (2) to explore the population dynamics.

```{r}
#for (each of these values) {
#    do the commands between the curly brackets
#}
```

In the loop you iterate the model through indexing your way in a vector (basically that you can extract the "ith"-value of a vector using hard brackets "[ ]"): 
```{r}
vec <- seq(1, 20, 1)
vec
vec[3] # the third element in vec is 3
```
... So we can build a for loop to simulate the time-dynamics in population density (N):
```{r}
# Create an empty vector to store the values in
N <- rep(NA, t) # this repeats the value NA t times
N[1] <- N_ini   # Set element number 1 to N_ini

for (i in 2:t) { # this performs the calculation (the model) in the loop from 2 to t, i is the current value
  N[i] <- r * N[i-1] * (1-(N[i - 1]/K)) + N[i-1]
}

N # look at what now is stored in N
```

```{r}
# Plot how population abundance change over time
plot(x = 1:t, y = N, xlab = "Time", ylab = "N", type = "l")
```

This plot doesn't show exactly how density dependence works in this model. To better understand that, we can plot the population growth rate as a function of $N$.  

```{r}
## We can start by creating a vector of N-values (ranging from 0 to carrying capacity):
N_vec <- seq(from = 0, to = K, by = 1)
N_vec

## Population growth rate
growth_pop <- r * N_vec * (1-(N_vec/K))

# Plot how population growth rate varies with population density
plot(growth_pop ~ N_vec, xlab = "N", ylab = "dN/dt", type = "l")
```
... and per capita growth rate as a function of $N$.
```{r}
## Per capita growth rate
growth_cap <- growth_pop / N_vec

# Plot how per capita growth rate varies with population density
plot(growth_cap ~ N_vec, xlab = "N", ylab = "dN/dt * 1/N", type = "l")
```

Use the above plots and your ecological knowledge to answer the following questions:

* **Question 1:** 
* a) How does density dependence operate (what does it do) in this model to the dynamics of the population and individual fish?
* b) What is the intercept and slope of the per capita growth rate?
* c) What is the main assumption we are making about individual fish when using the per capita growth rate to understand population growth in the Schaefer model?
* d) How could such density dependence occur in nature?
* e) At what density is the population growth rate highest? Show this analytically.
* f) What does carrying capacity represent in nature? Give examples from nature of why assuming a set carrying capacity is it wrong in a dynamic model.

* **Question 2:** 
* a) How would the population dynamics change if the intrinsic growth rate or carrying capacity would change? Explore this graphically by adding the updated models in the same plots you did above. Try with half of the initial parameter value **(r/2 and K/2)**.
* b) Give examples of species (not necessarliy fish) that matches the dynamics given by K and r of N, N2 and N3.
* c) what are the vertical lines in the plot below?

```{r}
# Create a new vector with new values of r and K
N2 <- rep(NA, t)
N2[1] <- N_ini
for (i in 2:150) {
N2[i] <- (r) * N2[i-1] * (1-(N2[i - 1]/K)) + N2[i-1]
}

# Create a new vector with new values of r and K
N3 <- rep(NA, t)
N3[1] <- N_ini
for (i in 2:t) {
N3[i] <- r * N3[i-1] * (1-(N3[i - 1]/(K))) + N3[i-1]
}

# To plot population abundance change over time: Plot the original plot and add the two lines by the command “lines”. Note that we have change the scale on the x axis to capture the asymptote for N2:
plot(y = N, x = 1:t, xlab = "Time", ylab = "N", type = "l", xlim =c(0,150) )
abline(h = K/2, lwd=1) # h = the y-intercept
lines(y = N2, x = 1:150, col="red")
abline(h = K/2, lty=4, lwd=1, col="red")
lines(y = N3, x = 1:t, col="green")
abline(h = K/4, col="green")
legend("bottomright", legend=c("N", "N2", "N3"),
       col=c("black","red", "green"), lty=1)
```
Try repeating this for the population growth rate and the per capita growth plots?

## Part 2: Equilibrium, harvesting and bifurcations
Now let us explore the equilibrium dynamics and dynamic stability of the Schaefer model analytically and graphically. Here asymptotic stability is an important concept. An equilibrium is asymptotically stable if the system returns to it after small deviations (perturbations) in density. The density at an equilibrium is denoted $N$*.

* **Question 3:** 
* a) What are the two equilibria of the Schaefer model? 
* b) Can you show these equilibria graphically? Assume catchability $q=0.2$ for the rest of this lab and effort $E=0.1$. 
* c) With this harvesting regime, what is the population density at the positive equilibrium? 

* **Question 5:** 
* a) What can we say about the dynamic stability of this equilibrium?
* b) By looking at the analytical solution for the non-trivial equilibrium, explain what happens with the equilibrium and its stability when we *gradually* increase the fishing effort? * c) show this graphically.

* **Question 7**: Plot $N$* as a function of fishing effort, $E$.

* **Question 8**: 
* a) Give two examples of what fishing effort can represents. 
* b) Can you plot yield, $Y=qEN$*, as a function of fishing effort, $E$?
* c) At what effort and density is long term yield maximized? 
* d) If our measure of effort represents number of boats, how will the yield per boat change with effort?
* e) How do the answers in b) and c) depend on $r$ and $K$? You can look at the analytical solutions and/or redo the last two plots.

* **Question 9**: Maximum sustainable yield (MSY) is the highest sustainable annual harvest rate (catch) at which the population biomass is sustained in the long term. What is the exact solution of the population density at MSY in the Schaefer model?

* **Question 10**: Solve for the yield at MSY in the Schaefer model.

## EXTRA BONUS part: Allee effects
If you feel that you are done with and understand question 1-10, continue with this voluntary exercise on Allee effects on population dynamics.

In the logistic model, per capita growth rates decreases linearly with population density. However, in many cases there is a threshold density for positive growth [2], i.e. an Allee effect. In other words, there is a positive relationship between per capita population growth rate and population size at low density. We can extend the Schaefer model to account for Allee-effects like this:
$$dN/dt = rN(N/K_0-1)(1-N/K)-qEN$$

* **Question 11**: What is the biological interpretation of the parameter $K_0$?

* **Question 12**: Plot the population- and per capita growth rate as a function of $N$ (you can set $K_0$=10). 

With this growth model we have an equilibrium at $N$*$=0$ (trivial equilibrium). The non-zero equilibria are solutions to the quadratic equation:
$$r(N/K_0-1)(1-N/K)=qE$$
which we obtain by equaling mortality and growth and dividing by $N$. Rather than solving this equation, we can explore it graphically for different fishing efforts the same way we did it before. Or, we can plot $N$ as a function of $qE$:

* **Question 13**: From the above equation, plot the equilibrium density $N$* as a function of $qE$. Compare to Question #6 and explain how and why the equilibria in the two cases differ! 

* **Question 14**: Plot yield as a function of fishing mortality (see Question #7). Locate the stable part of this curve. Which effort maximizes yield?

* **Question 15**: What challenges to fisheries management do Allee effects pose?

## References
[1] Schaefer, M. B. 1954. Some aspects of the dynamics of populations important to the management of commercial marine fisheries. *Bulletin of the Inter-American Tropical Tuna Commission*, 1, 25-56.

[2] Perälä, T., & Kuparinen, A. (2017). Detection of Allee effects in marine fishes: Analytical biases generated by data availability and model selection. *Proceedings of the Royal Society B*, 284(1861), 20171284.

[3] https://balticeye.org/en/fisheries/understanding-msy/


